---
title: "Multivariate Continuous Data"
author: "Prof. Joyce Robbins"
output: 
  slidy_presentation:
    css: styles5293.css
    fig_height: 4
    fig_width: 6
---

```{r, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                      cache = FALSE, echo = TRUE)
```

## `ggplot2`

If you're doing it right, legends appear automatically.

```{r legends, out.width = "50%"}
library(tidyverse)
setosa <- iris[iris$Species == "setosa",]
versicolor <- iris[iris$Species == "versicolor",]
virginica <- iris[iris$Species == "virginica",]

ggplot(setosa, aes(Sepal.Length, Sepal.Width)) +
  geom_point(color = "red") +
  geom_point(data = versicolor, aes(Sepal.Length, Sepal.Width),
             color = "green") +
  geom_point(data = virginica, aes(Sepal.Length, Sepal.Width),
             color = "blue")

ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point()
```

## `ggplot2`
```{r ref.label="legends", fig.height = 3, echo = TRUE}
```

## `ggplot2`

> - Do not use `grid.arrange`, `multiplot`, etc. (until further notice)

> - Use `facet_wrap()` or `facet_grid()` for multiple plots.

> - Watch out for function conflicts

> - Watch out for cache problems

## Parallel coordinate plots

> - `ggplot2::geom_line()` 

> - `GGally::ggparcoord()`   (static, ggplot2)

> - `MASS:: parcoord()`    (static, base)

> - `parcoords::parcoords()`  (interactive) <br> devtools::install_github("timelyportfolio/parcoords") <br> http://www.buildingwidgets.com/blog/2015/1/30/week-04-interactive-parallel-coordinates-1

> - ????


## Two continuous variables: scatterplot
```{r}
library(tidyverse)
theme_set(theme_grey(18))
norm20 <- rnorm(100, 20, 5)
norm5 <- rnorm(100, 5, 1)
df <- data.frame(norm20, norm5)
ggplot(df, aes(norm20, norm5)) + geom_point()
```

## Slope graph
```{r}
tidydf <- df %>% rownames_to_column("ID") %>% gather(var, value, -ID)
ggplot(tidydf, aes(x = var, y = value, group = ID)) + geom_line()
```

## Standardize data
```{r}
standardize <- function(x) (x-mean(x))/sd(x)
df2 <- tidydf %>% group_by(var) %>% 
  mutate(value = standardize(value)) %>% ungroup()
g0 <- ggplot(df2, aes(x = var, y = value, group = ID)) + 
  geom_line(alpha = .5) + ggtitle("Standardize")
g0  
```

## Rescale data to [0,1]
```{r}
df2 <- tidydf %>% group_by(var) %>% 
  mutate(value = scales::rescale(value)) %>% ungroup()
g1 <- ggplot(df2, aes(x = var, y = value, group = ID)) + 
  geom_line(alpha = .5) + ggtitle("Rescale")
g1
```

## Compare
```{r, out.width="50%", echo = FALSE}
g0
g1
```

## Compare different distributions
```{r}
notnorm <- c(1, rep(50, 98), 100)
df <- data.frame(norm20, notnorm)
tidydf <- df %>% rownames_to_column("ID") %>% gather(var, value, -ID)
ggplot(tidydf, aes(x = var, y = value, group = ID)) + 
  geom_line(alpha = .5)
```

## Compare
```{r, out.width= "50%"}
df2 <- tidydf %>% group_by(var) %>% 
  mutate(value = standardize(value)) %>% ungroup()
ggplot(df2, aes(x = var, y = value, group = ID)) + 
  geom_line(alpha = .5) + ggtitle("Standardize")
df2 <- tidydf %>% group_by(var) %>% 
  mutate(value = scales::rescale(value)) %>% ungroup()
ggplot(df2, aes(x = var, y = value, group = ID)) + 
  geom_line(alpha = .5) + ggtitle("Rescale")
```

## Scale = "globalminmax"
```{r, fig.width = 10}
library(GGally)
theme_set(theme_grey(14))
# scale = globalminmax
mystates <- data.frame(state.x77) %>% 
    rownames_to_column("State") %>% 
    mutate(Region = factor(state.region))
# state.region is a separate vector -- see: ?state
mystates$Region <- factor(mystates$Region,
                          levels = c("Northeast", "North Central",
                                     "South","West"))
ggparcoord(mystates, columns = 2:9, scale = "globalminmax")


```

## Scale = "std" (default) 
```{r, fig.width = 10}
# scale = std (default)
ggparcoord(mystates, columns = 2:9)
```

## Alpha
```{r, fig.width = 10}
# scale = std (default)
ggparcoord(mystates, columns = 2:9, alphaLines = .3)
```


## Alpha + rescale
```{r, fig.width = 10}
# scale = std (default)
ggparcoord(mystates, columns = 2:9, alphaLines = .3,
           scale = "uniminmax")
```

## Dataset with repeats
```{r}
x <- 1:10
y <- c(2,2,4,4,5,5,5,10,10,10)
z <- c(3,3,2,3,3,7,7,5,7,7)
w <- c(1, 1, 1, 7, 7, 7, 8, 8, 8, 8)
df <- data.frame(x,y,z, w)
g0 <- ggparcoord(df, columns = 1:4, scale = "globalminmax") + geom_vline(xintercept = 1:4, color = "lightblue")
g0
```

## Splines
```{r}
g1 <- ggparcoord(df, columns = 1:4, scale = "globalminmax",
                 splineFactor = 10) + 
  geom_vline(xintercept = 1:4, color = "lightblue")
g1
```

## Compare
```{r, out.width="50%", echo = FALSE}
g0
g1
```

## Alpha
```{r}
g1 <- ggparcoord(df, columns = 1:4, scale = "globalminmax",
                 splineFactor = 10, alphaLines = .5) + 
  geom_vline(xintercept = 1:4, color = "lightblue")
g1
```

## Alpha + rescale + splines
```{r, fig.width = 10}
# scale = std (default)
ggparcoord(mystates, columns = 2:9, alphaLines = .3,
           scale = "uniminmax", splineFactor = 10) + 
  geom_vline(xintercept = 2:9, color = "lightblue")

cat(colnames(mystates[2:9]))
```

## Watch out for categorical variables
```{r}
# See: http://www.buildingwidgets.com/blog/2015/1/30/week-04-interactive-parallel-coordinates-1
# devtools::install_github("timelyportfolio/parcoords")
library(parcoords)
data.frame(Titanic)  %>%
  parcoords(
    rownames = F # turn off rownames from the data.frame
    , brushMode = "1D-axes"
    , reorderable = T
    , queue = T
    , color = list(
      colorBy = "Class"
      ,colorScale = htmlwidgets::JS("d3.scale.category10()")
    )    
  )
```

## Add alpha blending
```{r}
# with alpha
mystates  %>% arrange(Region) %>% 
  parcoords(
    rownames = F 
    , brushMode = "1D-axes"
    , reorderable = T
    , queue = T
    , alpha = .5
    )    
```

## Color
```{r}
  parcoords(mystates
    , rownames = F 
    , brushMode = "1D-axes"
    , reorderable = T
    , queue = T
    , color = list(
      colorBy = "Region"
      ,colorScale = htmlwidgets::JS("d3.scale.category10()") 
    )    
  )
```

## Filter out Alaska / Texas
```{r}
# filter out Alaska, Texas, group by Region
myregionorder <- mystates %>% group_by(Region) %>% 
    summarize(n = n()) %>% arrange(n)
mystates$Region <- factor(mystates$Region, 
                          levels = myregionorder$Region)
data.frame(mystates)  %>% filter(State != "Alaska") %>%
  filter(State != "Texas") %>% arrange(Region) %>% 
  parcoords(
    rownames = F 
    , brushMode = "1D-axes"
    , reorderable = T
    , queue = T
    , color = list(
      colorBy = "Region"
      ,colorScale = htmlwidgets::JS("d3.scale.category10()") 
    )    
  )
```

## Highlighting a trend 
```{r}
mystates <- mystates %>% 
    mutate(color = factor(ifelse(Murder > 11, 1, 0))) %>% 
    arrange(color)
ggparcoord(mystates, columns = 2:9, groupColumn = "color") +
    scale_color_manual(values = c("grey70", "red")) +
    coord_flip() + guides(color = FALSE) +
    ggtitle("States with Murder Rate > 11 (per 100000) in red")
```

## 2017 Red Sox
```{r}
# Data from http://mlb.mlb.com/stats/
baseball <- read.csv("RedSoxStats2017.csv")
ggparcoord(baseball, columns = 5:20, groupColumn = "Pos",
           scale = "uniminmax")
```

## 2017 Red Sox (Interactive)
```{r}
baseball <- read.csv("RedSoxStats2017.csv")
myposorder <- baseball %>% group_by(Pos) %>% 
    summarize(n = n()) %>% arrange(n)
baseball$Pos <- factor(baseball$Pos, 
                          levels = myposorder$Pos)
data.frame(baseball)  %>% arrange(Pos) %>% 
    select(-Team) %>% 
  parcoords(
    rownames = F # turn off rownames from the data.frame
    , brushMode = "1D-axes"
    , reorderable = T
    , queue = T
    , color = list(
      colorBy = "Pos"
      ,colorScale = htmlwidgets::JS("d3.scale.category10()")
    )    
    , width = 1000
    , height = 400
  )
```

## 2016 Top Hitters (25 highest AVG)
```{r}
baseball <- read.csv("TopHitters2016.csv")
myteamorder <- baseball %>% group_by(Team) %>% 
    summarize(n = n()) %>% arrange(n)
baseball$Team <- factor(baseball$Team, 
                          levels = myteamorder$Team)
data.frame(baseball)  %>% arrange(Team) %>% 
  parcoords(
    rownames = F # turn off rownames from the data.frame
    , brushMode = "1D-axes"
    , reorderable = T
    , queue = T
    , color = list(
      colorBy = "Team"
      ,colorScale = htmlwidgets::JS("d3.scale.category10()")
    )    
    , width = 1000
    , height = 400
  )
```

## SAT data

https://data.cityofnewyork.us/Education/SAT-College-Board-2010-School-Level-Results/zt9s-n5aj/data

